Resistive magneto-rotational instability in massive protoplanetary disks

### 2013-10-06 paper/linear.tex
minor paper update

We now eliminate $\dd\bm{B}$ and $\dvz$ between the linearized
equations to obtain a system of ordinary differential equations for
$\bm{U}=\left(\dvx,\dvy,W,\dphi\right)$. We detail this technical
procedure in Appendix \ref{reduction}. The schematic problem is
\begin{align}
  L_{11}\dvx + L_{12}\dvy + L_{13}W + L_{14}\dphi &= 0, \label{lin1}\\
  L_{21}\dvx + L_{22}\dvy + L_{23}W + L_{24}\dphi &= 0, \label{lin2}\\
  L_{31}\dvx \quad\quad\quad\,\,\,\,\,\,\, + L_{33}W + L_{34}\dphi &=0,\label{lin3}\\
  \phantom{L_{31}\dvx + L_{22}\dvy +} L_{43}W + L_{44}\dphi
  &=0\label{lin4}, 
\end{align}
where the differential operators $L_{1j}$, $L_{2j}$ and $L_{3j}$ can
be read off Eq. \ref{final_vx}, Eq. \ref{final_vy} and 
Eq. \ref{final_w} respectively, and $L_{4j}$ from
Eq. \ref{lin_poisson}.

Note that $L_{31}\propto k_x$, so that if $k_x=0$ 
then Eq. \ref{lin1}---\ref{lin2} are decoupled from 
Eq. \ref{lin3}---\ref{lin4}.

@@

\subsection{Numerical procedure}
We use a pseudo-spectral approach to solve the set of linearized
equations. Let
\begin{align}\label{cheby_expand}
  \bm{U}(z) 
  = \sum_{k=1}^{N_z} \bm{U}_k\psi_k(z/\zmax), 
\end{align}
where $  \psi_k  = T_{2(k-1)},  $ and  $T_l$ is a Chebyshev polynomial of the first
kind of order $l$ \citep{stegun65}. %The number of basis functions is
%$N_z$ and $l_\mathrm{max}=2(N_Z-1)$ is the highest polynomial order. 
Note that we only use even Chebyshev polynomials which automatically satisifies the symmetry 
condition at $z=0$. Henceforth we only consider $z\geq 0$. 
The pseudo-spectral coefficients $\bm{U}_n$  are obtained by demanding
the set of linear equations to be satisfied at $N_z$ collocation
points along the vertical direction. We choose these points as the
extrema of $T_{2(N_z-1)}$ plus end points.

@@

\subsection{Diagnostics}

>>>>

We now eliminate $\dd\bm{B}$ and $\dvz$ between the linearized
equations to obtain a system of ordinary differential equations for
$\bm{U}=\left(\dvx,\dvy,W,\dphi\right)$. We detail the steps 
in Appendix \ref{reduction}. The problem is schematically 
\begin{align}
  L_{11}\dvx + L_{12}\dvy + L_{13}W + L_{14}\dphi &= 0, \label{lin1}\\
  L_{21}\dvx + L_{22}\dvy + L_{23}W + L_{24}\dphi &= 0, \label{lin2}\\
  L_{31}\dvx \quad\quad\quad\,\,\,\,\,\,\, + L_{33}W + L_{34}\dphi &=0,\label{lin3}\\
  \phantom{L_{31}\dvx + L_{22}\dvy +} L_{43}W + L_{44}\dphi
  &=0\label{lin4}, 
\end{align}
where the differential operators $L_{1j}$, $L_{2j}$ and $L_{3j}$ can
be read off Eq. \ref{final_vx}, Eq. \ref{final_vy} and 
Eq. \ref{final_w} respectively, and $L_{4j}$ from
Eq. \ref{lin_poisson}.

Note that $L_{31}\propto k_x$, so for $k_x=0$ 
Eq. \ref{lin1}---\ref{lin2} are decoupled from 
Eq. \ref{lin3}---\ref{lin4}. In this case, a solution is $\dphi=W=0$,
with $\dvx$ and $\dvy$ determined by Eq. \ref{lin1}---\ref{lin2},
which is the incompressible MRI problem. This means that the density
and potential perturbations can only influence the MRI for
$k_x\neq0$. However, SG will still have an effect through the
background equilibrium for $k_x=0$.

@@

\subsection{Numerical procedure}
We use a pseudo-spectral approach to solve the set of linearized
equations. Let
\begin{align}\label{cheby_expand}
  \bm{U}(z) 
  = \sum_{k=1}^{N_z} \bm{U}_k\psi_k(z/\zmax), 
\end{align}
where $  \psi_k  = T_{2(k-1)},  $ and  $T_l$ is a Chebyshev polynomial of the first
kind of order $l$ \citep{stegun65}. %The number of basis functions is
%$N_z$ and $l_\mathrm{max}=2(N_Z-1)$ is the highest polynomial order. 
Note that we only use even Chebyshev polynomials so the symmetry 
condition at $z=0$ is satisfied. Henceforth we only consider $z\geq 0$. 
The pseudo-spectral coefficients $\bm{U}_n$  are obtained by demanding
the set of linear equations to be satisfied at $N_z$ collocation
points along the vertical direction. We choose these points as the
extrema of $T_{2(N_z-1)}$ plus end points.

@@

\subsection{Diagnostics}
We visualize results in terms of energy densities. We define
\begin{align}
  &E_1 \equiv \frac{|\delta \bm{B}|^2}{2\beta},\\
  &E_2 = \frac{\rho}{\hat{c}_s^2}|\real\left(W\dphi^*\right)|,\\
  &E_3 = \frac{1}{2}\rho|\delta\bm{v}|^2,\\
  &E_4 = \frac{\rho|W|^2}{2\hat{c}_s^2},
\end{align}
as the perturbed magnetic, gravitational, kinetic and thermal
energies, respectively, which are functions of $z$. We
then normalize each by $\mathrm{max}(E_1)$.

### 2013-10-06 paper/setup.tex
minor paper update

We solve for $\hat{\rho}\equiv\rho/\rho_0$ subject to $\hat{\rho}=1$
and $d\hat{\rho}/dz=0$  at $z=0$. This is done numerically for
isothermal disks and analytically for the polytropic disk (see
Appendix \ref{appen1}). Examples of density profiles are shown in
Fig. \ref{eqm_den}.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_iso_density}
  \includegraphics[width=\linewidth]{figures/compare_poly_density}
  \caption{caption\label{eqm_den}}
\end{figure}

\subsection{Resistivity profile}
Following \cite{fleming03}, we adopt the resistivity profile
\begin{align}
  \eta(z) =
  \sqrt{2}\eta_0\left[\exp{\left(-g_+\right)}+\exp{\left(-g_-\right)}\right]^{-1/2},  
\end{align}
where
\begin{align}
  &g_\pm(z) =  \frac{\Sigma_\pm(z)-\Sigma_0}{\Sigma_*}, \\
  &\Sigma_\pm(z) = \int_{\pm z}^\infty\rho(z^\prime)dz^\prime, \label{sigma_pm}
\end{align}
and $\Sigma_0\equiv\Sigma_{\pm}(0)$, so that $g_\pm(0)=0$ and $\eta_0 
= \eta(0)$. The constant $\Sigma_*$ is chosen such that 
\begin{align}
  \cosh{\left(\frac{\Sigma_0}{\Sigma_*}\right)} =
  \left[\frac{\eta_0}{\eta(\infty)}\right]^2,
\end{align}
and we define $\eta_0/\eta(\infty)\equiv A_\eta$ as the conductivity
boost factor from the midplane to the disk surface. We remark that
once the equilibrium $\rho$ and $d\rho/dz$ are obtained from
Eq. \ref{eqm_eqns1} --- \ref{eqm_eqns2}, the integration for
Eq. \ref{sigma_pm} can be done implicitly by using the Poisson
equation.

>>>>

We solve for $\hat{\rho}\equiv\rho/\rho_0$ subject to $\hat{\rho}=1$
and $d\hat{\rho}/dz=0$  at $z=0$. This is done numerically for
isothermal disks and analytically for the polytropic disk (see
Appendix \ref{appen1}). Examples of density profiles are shown in
Fig. \ref{eqm_den}. The isothermal disk shows that self-gravity
compresses the disk and reduces the true pressure scale-height. The
polytropic disk shows little difference between different strengths of
self-gravity because the vertical co-ordiante is already normalized by
the disk thickness.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_iso_density}
  \includegraphics[width=\linewidth]{figures/compare_poly_density}
  \caption{Equilibrium density field from solving Eq. \ref{eqm_eqns1}
    --- \ref{eqm_eqns2} subject to an isothermal (top) and polytropic
    (bottom) equation of state. Note that the vertical co-ordinate is
    normalized differently for convenience. In the isothermal disk,
    $H$ is the characteristic scale-height in the
    non-self-gravitating limit; in the polytropic disk $H$ is the
    disk thickness. \label{eqm_den}}
\end{figure}

\subsection{Resistivity profile}
Following \cite{fleming03}, we adopt the resistivity profile
\begin{align}
  \eta(z) =
  \sqrt{2}\eta_0\left[\exp{\left(-g_+\right)}+\exp{\left(-g_-\right)}\right]^{-1/2},  
\end{align}
where
\begin{align}
  &g_\pm(z) =  \frac{\Sigma_\pm(z)-\Sigma_0}{\Sigma_*}, \\
  &\Sigma_\pm(z) = \int_{\pm z}^\infty\rho(z^\prime)dz^\prime, \label{sigma_pm}
\end{align}
and $\Sigma_0\equiv\Sigma_{\pm}(0)$, so that $g_\pm(0)=0$ and $\eta_0 
= \eta(0)$. The constant $\Sigma_*$ is chosen such that 
\begin{align}
  \cosh{\left(\frac{\Sigma_0}{\Sigma_*}\right)} =
  \left[\frac{\eta_0}{\eta(\infty)}\right]^2,
\end{align}
and we define $\eta_0/\eta(\infty)\equiv A$ as the conductivity
boost factor from the midplane to the disk surface. We remark that
once the equilibrium $\rho$ and $d\rho/dz$ are obtained from
Eq. \ref{eqm_eqns1} --- \ref{eqm_eqns2}, the integration for
Eq. \ref{sigma_pm} can be done implicitly by using the Poisson
equation.
